---
title: "Project 1 Final Deliverable"
author: "Faran Igani and Tyler Stoen"
format:
  html:
    theme: cosmo
    embed-resources: true
    code-fold: true
    code-summary: "Show the code"
editor: visual
code-tools: true
execute:
  warning: false
  message: false
---

## Packages and Data

```{r}
#| code-fold: true
#| label: Load packages 
library(tidyverse)
library(gt)
library(rvest)
library(ggplot2)
library(dplyr)
library(httr2)
library(jsonlite)
library(httr)
library(glue)
library(tidyjson)
library(ggridges)
library(scales)
library(stringdist)
library(tidytext)
```

## Github URL

<https://github.com/FaranIgani/STAT-541---Project-1>

# Data Context

```{r}
#| label: Load the Data
plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')
```

This dataset comes from global plastic waste cleanup efforts organized by [Break Free From Plastic]{.underline}, where volunteers collect and categorize plastic items found in the environment through “brand audits.” The cleaned version of the data was compiled and published by [Sarah Sauvé]{.underline}, who documented her approach in a blog post and shared the dataset to raise awareness and support broader efforts to understand plastic pollution sources. The data were collected across multiple countries worldwide, and each row represents the outcome of a cleanup for a specific country, year (2019 or 2020), and parent company, which serves as the identified source of the plastic. The dataset records counts of different types of plastic materials, including HDPE (high-density polyethylene), LDPE (low-density polyethylene), PET (polyester), polypropylene, polystyrene, PVC, and other, with a grand total variable representing the sum of all plastic items collected. These data were gathered during organized cleanup events in which volunteers manually collected waste and categorized items by material type and branding, allowing for company-level attribution when possible.

With this dataset, we are able to look and study how plastic waste varies by geography, material type and source company across a two year period and to better understand the sources and composition of plastic pollution globally. By identifying which companies contribute most to plastic waste and how waste varies across countries and materials, the dataset supports efforts to raise awareness, inform policy, and encourage corporate accountability. The data are of particular interest because plastic pollution is a major environmental issue, and understanding its sources is a critical step toward reducing its impact.

The data was cleaned by first combining multiple country-level CSV files for each year into a single dataset using row-binding. Column types were explicitly specified to ensure consistency across files, with categorical variables stored as character and count variables as numeric. A year variable was added to distinguish observations from 2019 and 2020. The grand total variable was converted from character to numeric using a parsing function, parse_number(), to remove punctuation. Names were also converted to lowercase. The 2019 data required additional cleaning to resolve inconsistencies, including extracting country names from file names, renaming variables to match the 2020 structure, and combining duplicated columns (e.g., pp and pp_2, ps and ps_2). After merging duplicate columns, the redundant variables were removed. These steps resulted in a unified dataset with consistent structure across both years.

On top of the data cleaning done as preparation, there were some observations with missing data. In many of the cases, plastic was all classified under "other", which provides an incorrect split of plastic types. This proves problematic for visualizing things like the split of plastic types because these plastics are almost certainly classified incorrectly. It would be misleading to replace these with zeros when investigating different plastic types, so instead we likely have to drop these rows to maintain correctness.

## Data from API

### API 1: restcountries.com

The additional country-level data were retrieved from the REST Countries API, which is a free, community-maintained RESTful API that aggregates geographic and demographic information for over 250 countries from various public sources. The data represent country-level attributes, so the observational unit is a country, with variables such as region, subregion, United Nations membership status, population density, and geographic coordinates. Unlike the plastics dataset, these data are not tied to a specific year; instead, they reflect the current state of country metadata at the time of access, as the API is continuously updated. The data were collected programmatically by sending an HTTP request to the API endpoint using the httr package in R, and the returned JSON data were parsed and converted into a tidy format using jsonlite. We chose to include this dataset to provide additional geographic and geo-political context for the countries in the plastic pollution data, allowing us to group, compare, and visualize patterns in plastic waste collection across regions and levels of development.

```{r}
#| label: data-from-api

base_url <- "https://restcountries.com/v4/"

restcountries <- GET(glue("{base_url}all?fields=name,continents,region,subregion,unMember,gini,density,geolocation"))

restcountries

restcountries_tidy <- restcountries %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>%
  as_tibble()

```

```{r}
#| label: merge-plastics

library(stringi)

normalize_name <- function(x) {
  x %>% 
    stri_trans_general("Latin-ASCII") %>% 
    str_to_lower() %>% 
    str_remove_all("[^a-z ]") %>% 
    str_trim() %>% 
    str_squish()
}

plastics_clean <- plastics %>% 
  filter(country != "EMPTY") %>% 
  mutate(
    country = case_when(
      country == "Korea" ~ "South Korea",
      country == "Taiwan_ Republic of China (ROC)" ~ "Taiwan",
      country == "United Kingdom of Great Britain & Northern Ireland" ~ 
        "United Kingdom",
      TRUE ~ country
    )
  ) %>%
  mutate(country_key = normalize_name(country))

restcountries_clean <- restcountries_tidy %>% 
  select(name.common, name.official, region, subregion,
         density, unMember, geolocation.latitude, 
         geolocation.longitude) %>% 
  mutate(common_key = recode(
    normalize_name(name.common),
    "ivory coast" = "cote divoire"
    ),
         official_key = normalize_name(name.official))
  

# Different name types, try both

joined <- plastics_clean %>%
  left_join(
    restcountries_clean %>% select(-official_key),
    by = c("country_key" = "common_key")
  )

matched <- joined %>% filter(!is.na(region))
unmatched <- joined %>% filter(is.na(region)) %>%
  select(names(plastics_clean))

refilled <- unmatched %>%
  left_join(
    restcountries_clean %>% select(-common_key),
    by = c("country_key" = "official_key")
  )

joined1 <- bind_rows(matched, refilled)
```

```{r}
#| label: merge-plastics-cont
## Fuzzy matching for parent companies using stringdist

noise_words <- c(
  "\\bthe\\b",
  "\\binc\\.?\\b",
  "\\bco\\.?\\b",
  "\\bltd\\.?\\b",
  "\\bcorp\\.?\\b",
  "\\bcompany\\b",
  "\\bcompanies\\b",
  "\\bgroup\\b",       # "Nestlé Group" -> "nestlé"
  "\\bholdings\\.?\\b",
  "\\bllc\\.?\\b",
  "\\bplc\\.?\\b",
  "\\bgmbh\\b",        # common German suffix
  "\\bsa\\b",          # common in Latin America/Europe
  "\\bsrl\\b"
)

normalize_company <- function(x) {
  x %>%
    str_to_lower() %>%
    str_remove_all(paste(noise_words, collapse = "|")) %>%
    str_remove_all("[^a-z0-9 ]") %>%
    str_squish()
}

resolve_typos <- function(names, threshold = 0.1) {
  dist_matrix <- stringdistmatrix(names, names, method = "jw")
  diag(dist_matrix) <- Inf
  
  canonical <- names
  
  for (i in seq_along(names)) {
    best_dist <- min(dist_matrix[i, ])
    if (best_dist < threshold) {
      best_match <- which.min(dist_matrix[i, ])
      canonical[i] <- min(names[i], names[best_match])
    }
  }
  
  tibble(normalized = names, canonical = str_to_title(canonical))
}

companies_raw <- unique(joined$parent_company)

name_map <- tibble(original = companies_raw) %>%
  mutate(normalized = normalize_company(original)) %>%
  left_join(
    resolve_typos(unique(.$normalized)),  # fuzzy match on normalized names only
    by = "normalized"
  )

canonical_labels <- name_map %>%
  count(canonical, original) %>%          # count how often each original appears
  slice_max(n, by = canonical, n = 1, with_ties = FALSE) %>% # pick the most frequent per canonical
  select(canonical, display_name = original)

joined <- joined1 %>%
  left_join(name_map %>% select(original, canonical),
            by = c("parent_company" = "original")) %>%
  left_join(canonical_labels, by = "canonical") %>% 
  mutate(parent_company = display_name) %>% 
  select(-display_name, -canonical)
```

### API 2: worldbank

We retrieved total population data and GDP data (in current U.S. dollars) for the years corresponding to collection years in the plastics dataset (2019 and 2020) from the World Bank’s World Development Indicators (WDI), a comprehensive collection of country-level development statistics compiled from official international sources. The data were published by the World Bank, with GDP and population estimates produced by World Bank country economists using information from national statistical agencies, the OECD, and other official country sources. The observational unit for this dataset is a country–year combination, matching the structure of the plastics data. The data correspond specifically to the years 2019 and 2020, aligning with the time frame of the cleanup records. These data were collected programmatically by sending requests to the World Bank Indicators REST API, and the returned JSON data were parsed and converted into a tidy format. We chose to include this dataset because it provides important economic context for the countries where plastic waste was collected, allowing us to explore how factors such as GDP and population relate to patterns in plastic waste across countries.

```{r}
#| label: API, World Bank  API

base_url <- "https://api.worldbank.org/v2/country/all/indicator/"

pop_code <- "SP.POP.TOTL"
gdp_code <- "NY.GDP.MKTP.CD"

pop_url <- str_c(base_url, pop_code, "?format=json&per_page=20000")
gdp_url <- str_c(base_url, gdp_code, "?format=json&per_page=20000")

pop_raw <- GET(pop_url) %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>%
  pluck(2) %>% 
  as_tibble() %>% 
  select(
    country.value,
    year = date,
    population = value
  ) %>% 
  mutate(year = as.numeric(year)) %>% 
  filter(year %in% c(2019, 2020))

gdp_raw <- GET(gdp_url) %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>% 
  pluck(2) %>% 
  as_tibble() %>% 
  select(
    country.value,
    year = date,
    gdp = value
  ) %>% 
  mutate(year = as.numeric(year)) %>% 
  filter(year %in% c(2019, 2020))

pop_and_gdp_raw <- inner_join(pop_raw, gdp_raw, by = c("country.value", "year")) %>% 
  mutate(country.value = recode(
           country.value,
          "Turkiye" = "Turkey",
          "Hong Kong SAR, China" = "Hong Kong",
          "Viet Nam" = "Vietnam",
          "Korea, Rep." = "South Korea",
          "United States" = "United States of America"
          ),
         country_code = normalize_name(country.value))

joined_all <- joined %>% 
  left_join(pop_and_gdp_raw, by = c("country_key" = "country_code", "year")) %>% 
  mutate(
    gdppercapita = gdp / population
  )
                              
```

# Research Questions

-   What are the characteristics of recorded plastic waste, in terms of the companies most responsible for the waste and the plastic types most commonly collected?

-   How does country geography and economic status effect their amount of recorded plastic waste? Do more developed countries in terms of GDP and population tend to have more or less recorded waste?

# Data Summaries & Visualizations

## **Research Question 1**

**Table 1**

```{r}
top_company_by_country <- function(ctry) {
  
  ctry_norm <- normalize_name(ctry)
  
  country_companies <- joined_all %>% 
    filter(country_key == ctry_norm) %>% 
    mutate(parent_company = str_squish(parent_company)) %>% 
    filter(!is.na(parent_company),
           !(parent_company %in% c("Grand Total", "null", "Unbranded", "Assorted", "NULL")))
  
  country_total <- country_companies %>% 
  summarise(total = sum(grand_total, na.rm = TRUE)) %>% 
  pull(total)
  
  country_companies %>% 
    group_by(parent_company) %>% 
    summarise(total = sum(grand_total, na.rm = TRUE)) %>% 
    slice_max(total, n = 1) %>% 
    mutate(country = str_to_title(ctry_norm),
           percent = total / country_total)
}

joined_all %>%
  distinct(country_key) %>%
  pull(country_key) %>% 
  map(top_company_by_country) %>% 
  list_rbind() %>% 
  filter(!is.nan(percent)) %>%
  select(country, parent_company, percent) %>% 
  arrange(desc(percent)) %>% 
  gt() %>% 
  cols_label(
    country        = "Country",
    parent_company = "Parent Company",
    percent        = "Percent"
  ) %>% 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) %>% 
  fmt_percent(columns = percent, decimals = 2) %>% 
  tab_header(
    title = md("**Company's Contributing
                 the Most Recorded Waste by Country**")
  ) %>% 
  tab_source_note(
    source_note = "Parent company with the largest number of recorded waste items in each country." 
  )
```

```{r}
#| label: Top companies by region
#| fig-alt: "Faceted bar chart showing the top five parent companies contributing to plastic waste within each region (Africa, Americas, Asia, and Europe). In all regions, 'Unbranded' plastic dominates by a large margin, especially in Asia where it far exceeds other companies. Named companies such as The Coca-Cola Company, Nestle, and Unilever appear across multiple regions but contribute substantially less than unbranded sources. Overall, the distribution is highly skewed, with a single dominant category in each region and relatively small contributions from the remaining companies."
#| fig-cap: "Bar chart showing the parent companies with the highest total recorded plastic waste, grouped by region."

top_companies_region <- joined_all %>%
  filter(
    !is.na(parent_company),
    !is.na(region),
    region != "Oceania",
    !(parent_company %in% c("Grand Total", "null", "Unbranded", "Assorted", "NULL"))
  ) %>%
  group_by(region, parent_company) %>%
  summarise(
    total_plastic = sum(grand_total, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(region) %>%
  slice_max(order_by = total_plastic, n = 5, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    company_ordered = reorder_within(parent_company, total_plastic, region)
  )

ggplot(top_companies_region,
       aes(x = total_plastic, y = company_ordered, fill = region)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ region, scales = "free") +
  scale_y_reordered() +
  scale_x_continuous(
  limits = c(0, 16000), # take out to not scale
  breaks = seq(0, 15000, by = 5000), # take out 
  labels = label_number(scale = 1e-3, suffix = "K")
) + 
  labs(
    title = "Top Parent Companies in Recorded Plastic Waste by Region",
    subtitle = "Top 5 companies with the highest recorded plastic waste in each region",
    x = "Total Plastic Collected",
    y = NULL
  ) +
  scale_fill_brewer(palette = "Dark2") +
  theme_light() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11),
    panel.grid.minor = element_blank()
  )
```

**Table 2**

```{r}
plastic_types <- c("hdpe", "ldpe", "o", "pet", "pp", "ps", "pvc")

dominant_plastic <- function(ctry) {
  
  ctry_norm <- normalize_name(ctry)
  
  df <- joined_all %>%
    filter(country_key == ctry_norm) %>%
    summarise(across(all_of(plastic_types), ~ sum(.x, na.rm = TRUE)))
  
  tibble(
    country       = str_to_title(ctry_norm),
    dominant_type = plastic_types[which.max(df)],
    dominant_n    = max(df)
  )
}

top_countries <- joined_all %>% 
  group_by(country_key) %>%
  summarise(total = sum(grand_total, na.rm = TRUE)) %>% 
  slice_max(total, n = 10) %>%
  pull(country_key)

top_countries %>%
  map(dominant_plastic) %>%
  bind_rows() %>%
  arrange(desc(dominant_n)) %>%
  mutate(dominant_type = dominant_type %>% 
           recode_values(
             "o" ~ "Other", 
             "ldpe" ~ "Low-Density Polyethylene",
             "hdpe" ~ "High-Density Polyethylene",
             "pet" ~ "Polyester",
             "pp" ~ "Polypropylene",
             "ps" ~ "Polystyrene",
             "pvc" ~ "PVC")
  ) %>% 
  gt() %>%
  tab_header(
    title = md("**Dominant Plastic Type for Top 10 Countries in Total Recorded Waste**")
    ) %>%
  tab_source_note(
    source_note = "Most commonly recorded plastic type for the countries producing the largest amount of plastic waste" 
  ) %>% 
  cols_label(
    country       = "Country",
    dominant_type = "Dominant Plastic Type",
    dominant_n    = "Amount of Dominant Plastic"
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) %>%
  fmt_number(columns = dominant_n, decimals = 0)
```

**Table 3**

```{r}
#| label: Modified custom function

calc_prop <- function(var, data) {
  data %>%
    filter(grand_total > 0, parent_company != "Grand Total") %>%
    group_by(year) %>%
    summarise(
      prop = sum(.data[[var]], na.rm = TRUE) / sum(grand_total, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(plastic_type = var)
}

plastic_vars <- c("empty", "hdpe", "ldpe", "o", "pet", "pp", "ps", "pvc")

table3_year <- map_dfr(plastic_vars, calc_prop, data = plastics) %>%
  mutate(
    plastic_type = recode(
      plastic_type,
      empty = "Empty",
      hdpe = "HDPE",
      ldpe = "LDPE",
      o = "Other",
      pet = "PET",
      pp = "PP",
      ps = "PS",
      pvc = "PVC"
    )
  ) %>%
  pivot_wider(
    names_from = plastic_type,
    values_from = prop
  ) %>%
  rename(Year = year)

table3 <- table3_year |>
  gt() |>
  cols_label(
    Year = md("**Year**"),
    Empty = md("**Empty (%)**"),
    HDPE = md("**HDPE (%)**"),
    LDPE = md("**LDPE (%)**"),
    Other = md("**Other (%)**"),
    PET = md("**PET (%)**"),
    PP = md("**PP (%)**"),
    PS = md("**PS (%)**"),
    PVC = md("**PVC (%)**")
  ) |>
  fmt_percent(
    columns = c(Empty, HDPE, LDPE, Other, PET, PP, PS, PVC),
    decimals = 1
  ) |>
  tab_header(
    title = md("**Proportion of Each Plastic Type by Year**")
  ) |>
  tab_source_note(
    source_note = "Percentages of each plastic type as a share of total plastic collected, by year."
  )

table3
```

**Figure 2**

```{r}
table3_long <- table3_year %>%
  pivot_longer(
    cols = -Year,
    names_to = "plastic_type",
    values_to = "prop"
  )

ggplot(table3_long, aes(x = prop, y = plastic_type, color = factor(Year))) +
  geom_segment(
    aes(x = 0, xend = prop, y = plastic_type, yend = plastic_type),
    linewidth = 1,
    alpha = 0.6
  ) +
  geom_point(size = 4) +
  scale_x_continuous(labels = percent_format()) +
  scale_color_brewer(palette = "Dark2", name = "Year") +
  labs(
    title = "Plastic Type Shares Differ Between 2019 and 2020",
    subtitle = "Lollipop chart comparing each plastic type as a proportion of total plastic collected",
    x = "Percent of Total Plastic",
    y = "Plastic Type"
  ) +
  theme_light(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

## **Research Question 2**

**Figure 1**

```{r}
#| fig-alt: "A ridgeline density plot showing the distribution of total plastic waste per country, separated by continent. Americas has the highest peak near zero with a moderate right tail. Asia has a flatter distribution with several secondary peaks at high waste values representing individual countries producing high waste. Africa is multimodal. Europe is concentrated near zero with one outlier peak. All distributions are right-skewed."
#| fig-cap: "Density ridge plot showing the distribution of total plastic collected across regions. Each ridge represents the distribution of country-level plastic totals within a region."

vis_1 <- joined_all %>%
  filter(!is.na(region), region != "Oceania") %>% 
  group_by(country, region) %>%
  summarize(
    total = sum(grand_total, na.rm = TRUE), .groups = "drop"
  ) %>% 
  mutate(
    region = fct_reorder(region, total, .desc = TRUE)
  ) %>% 
  ggplot(aes(x=total, y=region, fill=region)) +
  geom_density_ridges(alpha = 0.5, height = 1, position = "identity") +
  labs(
  title = "Total Plastic Waste by Continent",
  subtitle = "Distribution of Total Plastic Waste by Continent",
  x = "Total Plastic Waste (Number of Items)",
  y = ""
  ) +
  scale_x_continuous(labels = label_comma()) +
  scale_fill_brewer(palette = "Pastel1") +
  guides(fill = guide_legend(reverse = TRUE)) +
  theme_minimal() +
  theme(legend.position = "none")

vis_1
```

**Table 1**

```{r}
joined2 %>%
  group_by(country, year) %>%
  summarise(
    total_plastic = sum(grand_total, na.rm = TRUE),
    population = mean(population, na.rm = TRUE),
    plastic_per_person = total_plastic / population,
    .groups = "drop"
  ) %>%
  arrange(desc(plastic_per_person))%>%
  slice_head(n = 20) %>% 
  gt() %>% 
  cols_label(
    country = md("**Country**"),
    year = md("**Year**"),
    total_plastic = md("**Total Plastic**"),
    population = md("**Population**"),
    plastic_per_person = md("**Plastic per Person**")
  ) %>% 
  fmt_number(
    columns = c(total_plastic, population),
    decimals = 0
  ) %>% 
  fmt_number(
    columns = plastic_per_person,
    decimals = 5
  ) %>% 
  tab_header(
    title = md("**Top 20 Countries by Plastic Contributed per Person**")
  ) %>% 
  tab_source_note(
    source_note = "Plastic per person was calculated as total plastic collected divided by population."
  )
```

**Table 2**

```{r}
gdp_combo <- joined_all %>%
  filter(!is.na(gdp), !is.na(gdppercapita), !is.na(country)) %>%
  group_by(country, year) %>%
  summarise(
    gdp = mean(gdp, na.rm = TRUE),
    gdp_per_capita = mean(gdppercapita, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(gdp))

gdp_combo %>%
  gt() %>%
  cols_label(
    country = md("**Country**"),
    year = md("**Year**"),
    gdp = md("**GDP (USD)**"),
    gdp_per_capita = md("**GDP per Capita (USD)**")
  ) %>%
  fmt_currency(
    columns = gdp,
    currency = "USD",
    decimals = 0
  ) %>%
  fmt_currency(
    columns = gdp_per_capita,
    currency = "USD",
    decimals = 0
  ) %>%
  tab_header(
    title = md("**GDP and GDP per Capita by Country and Year**")
  ) %>%
  tab_source_note(
    source_note = "GDP and GDP per capita values are reported in current U.S. dollars from the World Bank."
  )
```

**Figure 1**

```{r}
#| fig-alt: "A ridgeline density plot showing the distribution of plastic waste per capita in pieces per person by continent, with four ridges for Europe, Americas, Asia, and Africa. All distributions are right-skewed and peak near zero plastic items recorded per person. Europe has the widest spread with several secondary peaks and one extreme outlier. Asia also has an extreme outlier at 0.010 pieces of plastic per capita. Africa is the most concentrated near zero with a slightly wider base than the others."
#| fig-cap: "Density ridge plot showing the distribution of pieces of plastic collected per capita across regions. Each ridge represents the distribution of country-level plastic totals per capita within a region."

vis_1.5 <- joined_all %>%
  filter(!is.na(region), region != "Oceania") %>% 
  group_by(country, region, population) %>%
  summarize(
    total = sum(grand_total, na.rm = TRUE), .groups = "drop"
  ) %>% 
  mutate(
    region = fct_reorder(region, total, .desc = TRUE),
    waste_per_capita = total/population
  ) %>% 
  ggplot(aes(x=waste_per_capita, y=region, fill=region)) +
  geom_density_ridges(alpha = 0.5, height = 1, position = "identity") +
  labs(
    title = "Plastic Waste per Capita by Continent",
    subtitle = "Distribution of Plastic Waste per Capita by Continent (in Pieces/Person)",
    x= "Plastic Waste Per Capita (Items per Person)",
    y= ""
  ) +
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal() +
  theme(legend.position = "none")

vis_1.5
```

**Figure 2**

```{r}
#| fig-alt: "A scatterplot comparing recorded plastic waste on the y-axis against GDP per capita on the x-axis, with points colored by continent. The majority of points are clustered near zero on both axes. One African country is a strong outlier with approximately 165,000 units of plastic waste recorded despite a very low GDP per capita below 5,000. Several Asian and Americas countries also show elevated waste counts at low GDP levels. European countries are spread across the full GDP per capita range but generally show low waste counts. No clear positive or negative relationship between GDP per capita and plastic waste is visible across the full dataset."
#| fig-cap: "Scatterplot showing the relationship between GDP per capita and total plastic collected across countries. Points are colored by region to highlight geographic differences."

vis_2 <- joined_all %>% 
  filter(!is.na(region), !is.na(gdppercapita)) %>%
  group_by(country, gdppercapita, region) %>% 
  summarize(
    total = sum(grand_total, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  ggplot(aes(x = gdppercapita, y = total, color = region)) +
  geom_point(alpha = 1, size = 1.5) +
  scale_x_continuous(labels = label_dollar()) +
  scale_color_brewer(
    palette = "Dark2",
    name = "World Region"
  ) +
  coord_cartesian(ylim = c(0, 180000)) +
  labs(
    title = "Higher-Income Countries Show Wide Variation in Plastic Waste",
    subtitle = "Country-level totals of recorded plastic waste vs GDP per capita",
    x = "GDP per Capita (USD)",
    y = "Total Plastic Waste"
  ) +
  theme_light(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

vis_2
```

# Data Model

```{r}
country_year <- joined_all %>%
  group_by(country, year, region) %>%
  summarise(
    total_plastic = sum(grand_total, na.rm = TRUE),
    gdp = mean(gdp, na.rm = TRUE),
    population = mean(population, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(
    !is.na(total_plastic),
    !is.na(gdp),
    !is.na(region)
  )

model_data <- country_year %>%
  filter(
    total_plastic > 0,
    !is.na(gdp),
   # !is.na(Temperature),
    !is.na(region)
  ) %>%
  mutate(
    log_total_plastic = log(total_plastic),
    log_gdp = log(gdp)
  )

plastic_model <- lm(
  log_total_plastic ~ log_gdp + region,
  data = model_data
)

summary(plastic_model)
```

**Model Table**

```{r}
library(broom)
library(gt)

model_table <- tidy(plastic_model, conf.int = TRUE) %>%
  select(term, estimate, std.error, statistic, p.value, conf.low, conf.high)

model_gt <- model_table %>%
  gt() %>%
  cols_label(
    term = md("**Term**"),
    estimate = md("**Estimate**"),
    std.error = md("**Std. Error**"),
    statistic = md("**t Statistic**"),
    p.value = md("**p-value**"),
    conf.low = md("**CI Lower**"),
    conf.high = md("**CI Upper**")
  ) %>%
  fmt_number(
    columns = c(estimate, std.error, statistic, conf.low, conf.high),
    decimals = 3
  ) %>%
  fmt_number(
    columns = p.value,
    decimals = 4
  ) %>%
  tab_header(
    title = md("**Regression Model for Log Total Plastic Collected**")
  ) %>%
  tab_source_note(
    source_note = "Linear regression model predicting log total plastic collected from GDP, population density, and region."
  )

model_gt
```

This model examines how economic factors (GDP), population density, and geographic region are associated with plastic waste. The coefficient for log GDP is positive (β = 0.271), suggesting that countries with higher economic output tend to have higher levels of plastic waste. This relationship is only marginally significant (p = 0.0619), however still moderate evidence of an association at the 5% level In contrast, population density does not appear to be a meaningful predictor (β = −0.030, p = 0.9147), suggesting that more densely populated countries do not necessarily generate more plastic waste after accounting for other variables.

Regional effects show clearer patterns. Compared to the reference region (Africa), countries in Asia (β = −1.662, p = 0.0469), Europe (β = −2.108, p = 0.0057), and Oceania (β = −4.601, p = 0.0344) have significantly lower levels of plastic waste, holding other variables constant. The Americas also show a negative relationship (β = −1.802), with moderate significant (p = 0.0587). These results suggest that geographic differences play an important role in plastic waste patterns, potentially reflecting differences in waste management systems, consumption behaviors, or environmental policies.

Overall, the model provides evidence that regional factors are more strongly associated with plastic waste than population density, while economic activity shows a weaker but suggestive relationship. This addresses our research questions by indicating that both economic and geographic context contribute to variation in plastic waste across countries, with region emerging as a key driver.